MLH1 Results Outline

### 2020-03-01 doc/Results.rmd
finalizing first draft

---
title: "MLH1 Results Outline"
date: 
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document: default
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
bibliography: refs_Main_Report.bib
---

@@

```{r mouse.var.plots.male, echo=FALSE, fig.width = 4, fig.height=2.5, include=FALSE, include=FALSE}

#variance plots
musc.male.VAR.plot <- ggplot(Musc.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Musc male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
musc.male.VAR.plot

Mol.male.VAR.plot <- ggplot(Mol.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Mol male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Mol.male.VAR.plot

Dom.male.VAR.plot <- ggplot(Dom.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Dom male ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Dom.male.VAR.plot
```

@@

lm.m2.Q12_male <- lm(var ~ strain, data= Q12_male_table)
lm.m2.Q12VAR.male_sum <- summary(lm.m2.Q12_male)
#round(lm.m2.Q12VAR.male_sum$coefficients[,c(1,4)], 5)
#NS
```

@@

@@

We applied a series of models to our dataset of mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co )` mice to fit the patterns of variation across sex, subspecies and strain into an evolutionary framework. All of the effects within the mixed model M1 were highly significant (LRT: subsp p= `r MLH1.MMlmer_results$sub.sp`, sex p = `r MLH1.MMlmer_results$sex.results`, subspecies * sex p = `r MLH1.MMlmer_results$interaction.pval`) and random LTR; strain p= `r MLH1.MMlmer_results$rand.pval`).

@@

```{r MLH1co.lmer.MM, echo=FALSE, warning = FALSE, digits = 3, message=FALSE}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

MLH1.MMlmer_results <- data.frame(sub.sp = MLH1.MM_subp, 
sex.results =  MLH1.MM_sex, 
               interaction.pval = MLH1.MM_interaction,  rand.pval = 
MLH1.MM_strain)
```

@@

```{r MLH1.glmM2, echo=FALSE, warning=FALSE,  digits = 3}

@@

```{r MLH1.glm.M3, echo=FALSE, digits = 3}

@@

The linear model results confirm the qualitative observations of sex-specific evolution and significant difference in the degree of sexual dimorphism in genome wide recombination rates. The subspecies fixed effects tested in M2, were either weak or not significant (glm; p=`r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]` and p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]` for _M. m. musculus_ and _M. m. molossinus_ respectively), indicating that most of the variance is due to strain effects. As such, we focus on results from M3. The two most significant strain effects were _domesticus^G^_ and _musculus^MSM^_ (p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]` for _domesticus^G^_ and _molossinus^MSM^_ respectively). In addition to significant strain effects for _molossinus^MSM^_, (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][5]` and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)  both _musculus^PWD^_ and _musculus^MSM^_ had significant strain by sex effects (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`
and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) _musculus^PWD^_ by male and _musculus^MSM^_ by male.

Combined, these results demonstrate that i) the significant sex-specific evolution in MSM and PWD and ii) the larger magnitude of heterochiasmy in _domesticus^G^_.

<The lack of a significant sex effect indicates that the evolution of mean MLH1 foci per cell is not uniform across sexes (but varies for specfic strains)>

@@

<!--
M2(glm; PWD*male p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][14]` and MSM-male p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][15]`)>
<Additionally, _domesticus^G^_ had significant strain and sex by strain effects (p =`r  summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]` for strain and strain by sex effect). These results confirm the qualtative pattern of the _domesticus^G^_ strain having the largest degree of sexual dimorphism (heterochiasmy) in the dataset.

@@

<SUMMERIZE BOTH THE MLH1 evolution and within mouse variance HERE>
<The greater amount of within mouse variance in females (almost and, (males have no detecable difference across strains for within mouse variance)
-- supports the distinct evolutionary pressures acting on the genome wide recombination rate in oocytes and spermatocytes.>

The striking patterns of variation across male and females suggest that the genome wide recombination rates have distinct evolutionary trajectories. For female the pattern of Strain averages distributed around a species wide average fits a model of stabilizing or relaxed evolution in contrast to the male pattern where there is rapid evolution in a subset of genetic strains that is close to a model of directional selection on genome wide recombination rate.

@@

#melt.Chrm.Prop.table = read.csv("MLH1repo/data/ChrmClassProportions2_melt.csv", header = TRUE)

melt.Chrm.Prop.table <- melt.Chrm.Prop.table[c(1:5)]

#add the subsps

melt.Chrm.Prop.table$strain<- factor(melt.Chrm.Prop.table$strain, levels =c("LEW",
                      "WSB","G", "TOM","AST","KAZ", "SKIVE","PWD","MSM", "MOLF"), order=T )

@@

chrm.class.barplot <- ggplot(data = melt.Chrm.Prop.table, aes(y = value, x = strain, fill=as.factor(type)) ) + geom_col(position = "dodge")+facet_grid(sex~strain, scales = "free")+ggtitle("Chrm Proportions from Hand Measured Data")+ylim(c(0,1))

@@

We distill the variance observed in the mean MLH1 foci counts per cell to number of MLH1 foci per bivalent. W use the term bivalent instead of chromosome because our observations of MLH1 foci are made of pachtyene cells, our chromosome observations are fully synapsed paired homologous chromosomes (4n copies of each chromosomes). Additionally we can not fully distinguish the identity of chromsomes (e.g. Chromosome 1 or Chromosome 2), for these reasons we use the 'term bivalent'.

@@

<Indicateing, SUMMARY the TRANSITION to single bivalen measures>

@@

<chromatin compaction background, process within cells - compaction ratio within cells>

@@

<t.test; p = `r Q1_mus.av_shortSC_t.test.WSB$p.value`, p =`r format(Q1_mus.av_shortSC_t.test.G$p.value, scientific = F)`, and p=`r format(Q1_mus.av_shortSC_t.test.LEW$p.value, scientific = F)` for _domesticus^WSB^_, _domesticus^G^_ and _domesticus^LEW^_ respectively. the musculus strins (t.test; p= `r format(Q1_mus.av_shortSC_t.test.PWD$p.value, scientific = F)`, `r Q1_mus.av_shortSC_t.test.KAZ$p.value`, and `r  format(Q1_mus.av_shortSC_t.test.SKIVE$p.value, scientific = F)` for _musculus^PWD^_ , _musculus^KAZ^_ , _musculus^SKIVE^_  respectively>

@@

```{r Q1.tot.SC_show, echo=FALSE, warning=FALSE, fig.width = 4, fig.height=2.5}

@@

#add category mean
#totSC.strain.av.table %>% filter
totsc.dom_tot.sc <- Q1.totSC.mouse.av.table %>% filter( subsp == "Dom" )

totsc_strain.av.dom <- totSC.strain.av.table %>% filter( subsp == "Dom" )

@@

  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+  ylim(c(1000,2600 ))+ggtitle("Mouse av. total SC area")

@@

Tot.SC_mouse.av_Mol
```

@@

<_domesticues^WSB^_ p = `r format(Q1_tot.sc_t.test.WSB$p.value, scientific = F )`, _domesticues^G^_  p = `r format(Q1_tot.sc_t.test.G$p.value, scientific= F )`, _domesticues^LEW^_ p = `r format(Q1_tot.sc_t.test.LEW$p.value, scientific= F )`, _musculus^PWD^_ p = `r format(Q1_tot.sc_t.test.PWD$p.value,scientific = F )`, _musculus^KAZ^_ p = `r format(Q1_tot.sc_t.test.KAZ$p.value, scientific = F )`, _molossinus^MSM^_ `r format(Q1_tot.sc_t.test.MSM$p.value, scientific = F )`>

**M1. Mixed Model** $$mouse \ average \ SC \ metric ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

<!--
In the mixed models for the sex differences mean **short bivalents** length as the dependent variable, sex is the most significant effect (LTR; p =`r Q1lmer.MM_SC_short_results$sex.results`). The interaction (subspecies by sex) effect was also slightly significant (p = `r Q1lmer.MM_SC_short_results$interaction.pval`).

@@

!! SUMARIZE and remove most of these pvalues!!

@@

For M3 in the mean total SC lengths there were many strain effects are significant, but the largest are sex (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][8]`)
but also some strain effects, G strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][2]`), LEW strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][3]`), PWD strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][4]`). (this means...). Supporting the strong sex difference and also indicating that SC length has evolved across strains within this panel, when chromatin condensation is summarized at this level. 
-->

<!--
(subsp*sex) mol.male  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][10]`, sex.g  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][11]`, lew.male p =`r sum.Q1.glm_totSC_M2$coefficients[,4][12]`
for the _mmdomesticues^WSB^ _, _mmdomesticues^G^ _, and _mmdomesticues^LEW^ _
-->

@@

#Q1. 1CO Position Sex Differences

@@

**M1. Mixed Model**

$$mouse \ average \ F1 position ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear model**

$$mouse \ average \ F1 position ~=~ subsp * sex * strain + \varepsilon $$

$$mouse \ average \ F1 position ~=~ sex * strain + \varepsilon $$

@@

For the Mixed model, sex is the most significant effect (LTR; p =`r lmer.Q1.nrm1CO_M1_results$sex.results`).The random effect of strain is also significant (LRT; p =`r lmer.Q1.nrm1CO_M1_results$rand.pval`).

The linear models were used to detect if specific strains differed in the sex differences in the 1CO landscape. While sex was the most significant effect for both M2 and M3 (there were some strain specific effects...)

@@

<!-- Ignore this section / kind of results for know
Chromosome size effects are a confounding factor for CO position, with the expectation that shorter chromosomes will have a more uniform landscape compared to larger/longer chromosomes (cite). To account for this we again use the reduced bivalent sets (long.biv and short.biv) -->

@@

#Q1. Sex Differences in CO Interference (IFD)

@@

<!-- lmer/ glms-->

Since the most significant t.test were significantly different across sexes for IFD^PER^ and we follow up with the model framework to.

To test for strain specific effects we applied the model framework

**M1. Mixed Model** $$mouse \ average \ IFD ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

$$mouse \ average \ IFD ~=~  sex * strain + \varepsilon $$

@@

For the Mixed models of IFDs, sex is a significant effect for both raw and nrmIFD.

(IFD.PER), sex is most significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`) subspecies, interaction are slightly significant. <ONLY REPORT FOR PER MEASURES>

For the normalized values in both M2 and M3, sex is a significant effect, increasing nrm.IFD in males. SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure (p = male-SKIVE p =  )

When interference is measured in physical SC units, the differences between sexes is low / only slightly significant -leading support that the phyisical measures of interference (in SC units) is conserved within species.

While when the variance in chromatin compaction is controlled males have stronger interference (like genetic maps) the impact on the rec landscapes being.

@@

Our results confirm the typical landscape (sex differences).

the terminal position of single foci and the greater distance between two foci on the same chromsome will have a consequence of

The genetic consequence of such a pattern is that larger 'chuncks' a greater proportion of linked sites will segregate together in males compared to females (Veller). (having the effect of fewer novel combinations of site pairs along the total chromosome).

@@

<pattern for the single bivalent, pos correlation, The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high recombiningstrains then low recombining strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low recombining strains, which pushes the mean SC length up. Where as in the high recombiningstrains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high recombininggroup. They have a lower probability of 2COs below a certain SC length threshold.>
```{r Q2.SC.short.long.setup, echo=FALSE}

@@

<!-- models -->

We follow up the general pattern of a weak signal of high recombiningstrains having longer SC lengths by running the model framework.

**Linear model M1 ** $$mouse \ average \ SC metric ~=~ subsp * strain + \varepsilon $$

**Linear model M2 ** $$mouse \ average \ SC metric ~=~ strain + \varepsilon $$

@@

For total SC in M1 (glm including subsp), the two subspecies effects are significant (glm; p= _Musculus_ `r sum.Q2.glm_mus.av_totSC_M1[,4][2]` and _Molossinus_ p= `r sum.Q2.glm_mus.av_totSC_M1[,4][3]`)

In the short bivalent data set, analyzed by M1, both subspecies are sig (glm; p= _Musculus_ `r sum.glmQ2_short.biv_M1$coefficients[,4][2]` and _Molossinus_ p= `r sum.glmQ2_short.biv_M1$coefficients[,4][3]`).

long bivalent set in M1 --

Musc `r sum.glmQ2_long.biv_M1$coefficients[,4][2]`

MSM strain `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`

<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).>

<!--
**M3 -- tests within subspecies!!** SO the M3 model can be run within species specific data -- to test the predictions (based on the variation in mean MLH1 counts), (use all 3 metircs) **the within subspecies M3 models don't fully meet the predictions
for total.Sc : there are significant Dom strain effects: in musc kaz is significant negative, several other strains have higher total SC, (MOL in the intercept is significant).
short biv: Dom: Significant G effect, MUSC no significant -- strain effects
long.biv:  Dom: no strain effects, MUSC no significant strain effects, MOL - not enough observations
**For the long bivalent dataset, In M1 where the subsp are tested, only Musc is significant (glm; Musc p= `r sum.glmQ2_long.biv_M1$coefficients[,4][2]`). The significant strain effects in M1 are slight just MSM (glm; MSM p= `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`).**

@@

#already made
#subset
male.bivdata.2CO_IFD <- Curated_BivData %>% filter(sex== "male") %>% filter(species =="M.musculus") %>% filter(hand.foci.count == 2)
#t.tests for the Rec group
Curated_BivData$Rec.group <- ifelse(grepl("PWD male", Curated_BivData$category), 1, 
                                    ifelse(grepl("MSM male", Curated_BivData$category), 1,
                                           ifelse(grepl("SKIVE male",Curated_BivData$category),1, 0)))

@@

The mouse averages for both IFD^raw^ and IFD^PER^ were significantly longer in the high recombining groups (t.test; p = `r t.test_mouse.av.IFD_PER$p.value` for IFD^PER^ and p = `r t.test_mouse.av.IFD_ABS$p.value` for IFD^raw^). The pattern is confirmed with t.test comparing groups within _musculus_ and _molossinus_
the same subepcies (t.tests p =`r t.test_mouse.av.IFD_PER_MUSC$p.value` and p= `r t.test_mouse.av.IFD_PER_MOL$p.value`).

insert the raw IFD t.test!!

@@

#M2.raw
Q2.M2glm_IFD.raw <- glm(mean_IFD.2CO_ABS ~ strain, data= mouse.avs_4MM %>% filter(sex == "male"))
#summary(Q2.M2glm_IFD.raw)

@@

```{r Q2.IFD_rate.shortIFD, echo=FALSE, warning = FALSE}

@@

<!-- final summary, BE brief

BE brief,

final summary, highlight the themes of 1. more variance in female traits (MLH1 foci per cell, rec landscape / interference)

while the male traits -- more focused (less variance in MLH1 counts per cell, shorter chromatin comparction?  bias position, stronger interference)

(should I measure the var of all traits in Q1)

2. rapid evolution of genome wide rec rates can fit into this -

- evolution of more COs requires more DSBs (evidence points to a earlier step which is likely the chromatin compaction (amount of axis))

>>>>

---
title: "MLH1 Results Outline"
date: '3/1/2020'
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document: default
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
bibliography: refs_Main_Report.bib
---

@@

```{r mouse.var.plots.male, echo=FALSE, fig.width = 4, fig.height=2.5, include=FALSE, include=FALSE, eval=FALSE}

#variance plots
musc.male.VAR.plot <- ggplot(Musc.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Musc male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Mol.male.VAR.plot <- ggplot(Mol.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Mol male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Dom.male.VAR.plot <- ggplot(Dom.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Dom male ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)

@@

lm.m2.Q12_male <- lm(var ~ strain, data= Q12_male_table)
lm.m2.Q12VAR.male_sum <- summary(lm.m2.Q12_male)
#round(lm.m2.Q12VAR.male_sum$coefficients[,c(1,4)], 5)
#NS

@@

```{r MLH1co.lmer.MM, echo=FALSE, waring=FALSE, message=FALSE}

#DF.HetC.MixedModel.HQ

MLH1.MMlmer_results <- data.frame(sub.sp = MLH1.MM_subp, 
sex.results =  MLH1.MM_sex, 
               interaction.pval = MLH1.MM_interaction,  rand.pval = 
MLH1.MM_strain)

@@

We applied a series of models to our dataset of mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice to fit the patterns of variation across sex, subspecies and strain into an evolutionary framework. All of the effects within the mixed model M1 were highly significant (LRT: subsp p= `r MLH1.MMlmer_results$sub.sp`, sex p = `r MLH1.MMlmer_results$sex.results`, subspecies * sex p = `r MLH1.MMlmer_results$interaction.pval`) and random LTR; strain p= `r MLH1.MMlmer_results$rand.pval`).

@@

@@

```{r MLH1.glmM2, echo=FALSE}

@@

```{r MLH1.glm.M3, echo=FALSE}

@@

The linear model results confirm the qualitative observations of sex-specific evolution and significant difference in the degree of sexual dimorphism in genome wide recombination rates. The subspecies fixed effects tested in M2, were either weak or not significant (glm; p=`r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]` and p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]` for _M. m. musculus_ and _M. m. molossinus_ respectively), indicating that most of the variance is due to strain effects. As such, we focus on results from M3. The two most significant strain effects were _domesticus^G^_ and _musculus^MSM^_ (p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]` for _domesticus^G^_ and _molossinus^MSM^_ respectively). In addition to significant strain effects for _molossinus^MSM^_, (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][5]` and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)  both _musculus^PWD^_ and _musculus^MSM^_ had significant strain by sex effects (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`
and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) _musculus^PWD^_ by male and _musculus^MSM^_ by male. Combined, these results demonstrate that i) the significant sex-specific evolution in MSM and PWD and ii) the larger magnitude of heterochiasmy in _domesticus^G^_.

@@

<!--

M2(glm; PWD*male p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][14]` and MSM-male p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][15]`)>
<Additionally, _domesticus^G^_ had significant strain and sex by strain effects (p =`r  summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]` for strain and strain by sex effect). These results confirm the qualtative pattern of the _domesticus^G^_ strain having the largest degree of sexual dimorphism (heterochiasmy) in the dataset.

@@

<SUMMERIZE BOTH THE MLH1 evolution and within mouse variance HERE>
<The greater amount of within mouse variance in females (almost and, (males have no detecable difference across strains for within mouse variance)
-- supports the distinct evolutionary pressures acting on the genome wide recombination rate in oocytes and spermatocytes.
The striking patterns of variation across male and females suggest that the genome wide recombination rates have distinct evolutionary trajectories. For female the pattern of Strain averages distributed around a species wide average fits a model of stabilizing or relaxed evolution in contrast to the male pattern where there is rapid evolution in a subset of genetic strains that is close to a model of directional selection on genome wide recombination rate.>

@@

#melt.Chrm.Prop.table = read.csv("MLH1repo/data/ChrmClassProportions2_melt.csv", header = TRUE)
melt.Chrm.Prop.table <- melt.Chrm.Prop.table[c(1:5)]

#add the subsps
melt.Chrm.Prop.table$strain<- factor(melt.Chrm.Prop.table$strain, levels =c("LEW",
                      "WSB","G", "TOM","AST","KAZ", "SKIVE","PWD","MSM", "MOLF"), order=T )

@@

chrm.class.barplot <- ggplot(data = melt.Chrm.Prop.table, aes(y = value, x = strain, fill=as.factor(type)) ) + geom_col(position = "dodge")+facet_grid(sex~strain, scales = "free")+ggtitle("Chrm Proportions from Hand Measured Data")+ylim(c(0,1))+theme(legend.position="none")

@@

We distill the variance observed in the mean MLH1 foci counts per cell to number of MLH1 foci per bivalent. Because our observations of MLH1 foci are made of pachtyene cells, our chromosome observations are fully synapsed paired homologous chromosomes (4n copies of each chromosomes). Additionally we can not fully distinguish the identity of chromsomes (e.g. Chromosome 1 or Chromosome 2), for these reasons we use the term 'bivalent' instead of chromosome.

@@

@@

@@

<!-- t.test; p = `r Q1_mus.av_shortSC_t.test.WSB$p.value`, p =`r format(Q1_mus.av_shortSC_t.test.G$p.value, scientific = F)`, and p=`r format(Q1_mus.av_shortSC_t.test.LEW$p.value, scientific = F)` for _domesticus^WSB^_, _domesticus^G^_ and _domesticus^LEW^_ respectively. the musculus strins (t.test; p= `r format(Q1_mus.av_shortSC_t.test.PWD$p.value, scientific = F)`, `r Q1_mus.av_shortSC_t.test.KAZ$p.value`, and `r  format(Q1_mus.av_shortSC_t.test.SKIVE$p.value, scientific = F)` for _musculus^PWD^_ , _musculus^KAZ^_ , _musculus^SKIVE^_  respectively -->

@@

```{r Q1.tot.SC_show, echo=FALSE, warning=FALSE}
# fig.width = 4, fig.height=2.5

@@

#add category mean
#totSC.strain.av.table %>% filter
#totsc.dom_tot.sc <- Q1.totSC.mouse.av.table %>% filter( subsp == "Dom" )
#totsc_strain.av.dom <- totSC.strain.av.table %>% filter( subsp == "Dom" )

@@

  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+  ylim(c(1000,2600 ))+ggtitle("Mouse av. total SC area") + theme(legend.position="none")

Tot.SC_mouse.av_ALL

@@

@@

<!-- _domesticues^WSB^_ p = `r format(Q1_tot.sc_t.test.WSB$p.value, scientific = F )`, _domesticues^G^_  p = `r format(Q1_tot.sc_t.test.G$p.value, scientific= F )`, _domesticues^LEW^_ p = `r format(Q1_tot.sc_t.test.LEW$p.value, scientific= F )`, _musculus^PWD^_ p = `r format(Q1_tot.sc_t.test.PWD$p.value,scientific = F )`, _musculus^KAZ^_ p = `r format(Q1_tot.sc_t.test.KAZ$p.value, scientific = F )`, _molossinus^MSM^_ `r format(Q1_tot.sc_t.test.MSM$p.value, scientific = F )` -->

@@

<!--
**M1. Mixed Model** $$mouse \ average \ SC \ metric ~=~ subsp * sex + rand(strain) + \varepsilon $$

In the mixed models for the sex differences mean **short bivalents** length as the dependent variable, sex is the most significant effect (LTR; p =`r Q1lmer.MM_SC_short_results$sex.results`). The interaction (subspecies by sex) effect was also slightly significant (p = `r Q1lmer.MM_SC_short_results$interaction.pval`).

@@

**!! SUMARIZE and remove most of the models pvalues!!**

@@

For M3 in the mean total SC lengths there were many strain effects are significant, but the largest are sex (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][8]`)
but also some strain effects, G strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][2]`), LEW strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][3]`), PWD strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][4]`). (this means...). Supporting the strong sex difference and also indicating that SC length has evolved across strains within this panel, when chromatin condensation is summarized at this level.

(subsp*sex) mol.male  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][10]`, sex.g  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][11]`, lew.male p =`r sum.Q1.glm_totSC_M2$coefficients[,4][12]`
for the _mmdomesticues^WSB^ _, _mmdomesticues^G^ _, and _mmdomesticues^LEW^ _
-->

@@

# Q1. 1CO Position Sex Differences

@@

@@

For the Mixed model, sex is the most significant effect (LTR; p =`r lmer.Q1.nrm1CO_M1_results$sex.results`).The random effect of strain is also significant (LRT; p =`r lmer.Q1.nrm1CO_M1_results$rand.pval`). The linear models were used to detect if specific strains differed in the sex differences in the 1CO landscape. While sex was the most significant effect for both M2 and M3 (there were some strain specific effects...)

@@

<!--
**M1. Mixed Model**
$$mouse \ average \ F1 position ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear model**
$$mouse \ average \ F1 position ~=~ subsp * sex * strain + \varepsilon $$

**M3. Linear model**
$$mouse \ average \ F1 position ~=~ sex * strain + \varepsilon $$

Ignore this section / kind of results for know
Chromosome size effects are a confounding factor for CO position, with the expectation that shorter chromosomes will have a more uniform landscape compared to larger/longer chromosomes (cite). To account for this we again use the reduced bivalent sets (long.biv and short.biv) -->

@@

# Q1. Sex Differences in CO Interference (IFD)

@@

<!-- lmer/ glms  **M1. Mixed Model** $$mouse \ average \ IFD ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

$$mouse \ average \ IFD ~=~  sex * strain + \varepsilon $$
-->

@@

Since the most significant t.test were significantly different across sexes for IFD^PER^ and we follow up with the model framework. For the Mixed models of IFDs, sex is a significant effect for both raw and nrmIFD. (IFD.PER), sex is most significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`) subspecies, interaction are slightly significant. <ONLY REPORT FOR PER MEASURES>.  For the normalized values in both M2 and M3, sex is a significant effect, increasing nrm.IFD in males. SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure (p = male-SKIVE p =  ).  When interference is measured in physical SC units, the differences between sexes is low / only slightly significant -leading support that the phyisical measures of interference (in SC units) is conserved within species. While when the variance in chromatin compaction is controlled males have stronger interference (like genetic maps) the impact on the rec landscapes being.

@@

Our results confirm the typical landscape (sex differences). The terminal position of single foci and the greater distance between two foci on the same chromsome will have a consequence of ... The genetic consequence of such a pattern is that larger 'chuncks' a greater proportion of linked sites will segregate together in males compared to females (Veller). (having the effect of fewer novel combinations of site pairs along the total chromosome).

@@

<pattern for the single bivalent, pos correlation, The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high recombiningstrains then low recombining strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low recombining strains, which pushes the mean SC length up. Where as in the high recombiningstrains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high recombininggroup. They have a lower probability of 2COs below a certain SC length threshold.>

```{r Q2.SC.short.long.setup, echo=FALSE}

@@

<!-- models 
**Linear model M1 ** $$mouse \ average \ SC metric ~=~ subsp * strain + \varepsilon $$
**Linear model M2 ** $$mouse \ average \ SC metric ~=~ strain + \varepsilon $$-->

@@

We follow up the general pattern of a weak signal of high recombiningstrains having longer SC lengths by running the model framework. For total SC in M1 (glm including subsp), the two subspecies effects are significant (glm; p= _Musculus_ `r sum.Q2.glm_mus.av_totSC_M1[,4][2]` and _Molossinus_ p= `r sum.Q2.glm_mus.av_totSC_M1[,4][3]`). (in the models using the short and long reduced datasets, (significant effects are /range from musc and strain MSM, (glm; p < 0.05). In the short bivalent data set, analyzed by M1, both subspecies are sig (glm; p= _Musculus_ `r sum.glmQ2_short.biv_M1$coefficients[,4][2]` and _Molossinus_ p= `r sum.glmQ2_short.biv_M1$coefficients[,4][3]`). While the models for the reduced dta set suggest that the subspecies with the high recombining strains have evolved longer SC lengths -- But the differences in // but the longer SC lengths do ot cleanily predict the higher recombining strains (there are longer SC in ). To some extent the SC lenth evolution is decoupled from the mean MLH1 foci per cell (since some strains have longer sc - but 'lower' recombination rates)

<!--
long bivalent set in M1 -- 
Musc `r sum.glmQ2_long.biv_M1$coefficients[,4][2]` 
MSM strain `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`
<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).>

**M3 -- tests within subspecies!!** SO the M3 model can be run within species specific data -- to test the predictions (based on the variation in mean MLH1 counts), (use all 3 metircs) **the within subspecies M3 models don't fully meet the predictions
for total.Sc : there are significant Dom strain effects: in musc kaz is significant negative, several other strains have higher total SC, (MOL in the intercept is significant).
short biv: Dom: Significant G effect, MUSC no significant -- strain effects
long.biv:  Dom: no strain effects, MUSC no significant strain effects, MOL - not enough observations
**For the long bivalent dataset, In M1 where the subsp are tested, only Musc is significant (glm; Musc p= `r sum.glmQ2_long.biv_M1$coefficients[,4][2]`). The significant strain effects in M1 are slight just MSM (glm; MSM p= `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`).**

@@

#ABS

t.test_mouse.av.IFD_ABS_MUSC <- t.test(  (mouse.avs_4MM %>% filter(Rec.group == "1") %>% filter(sex == "male") %>% filter(subsp == "Musc") )$mean_IFD.2CO_ABS,
   (mouse.avs_4MM %>% filter(Rec.group == "0") %>% filter(sex == "male")  %>% filter(subsp == "Musc"))$mean_IFD.2CO_ABS   )

t.test_mouse.av.IFD_ABS_MOL <- t.test(  (mouse.avs_4MM %>% filter(Rec.group == "1") %>% filter(sex == "male") %>% filter(subsp == "Mol") )$mean_IFD.2CO_ABS,
   (mouse.avs_4MM %>% filter(Rec.group == "0") %>% filter(sex == "male") %>% filter(subsp == "Mol") )$mean_IFD.2CO_ABS  )

@@

The mouse averages for both IFD^raw^ and IFD^PER^ were significantly longer in the high recombining groups (t.test; p = `r t.test_mouse.av.IFD_PER$p.value` for IFD^PER^ and p = `r t.test_mouse.av.IFD_ABS$p.value` for IFD^raw^). The pattern is confirmed with t.test comparing groups within _musculus_ and _molossinus_
the same subepcies (t.tests p =`r t.test_mouse.av.IFD_PER_MUSC$p.value` and p= `r t.test_mouse.av.IFD_PER_MOL$p.value` for IFD^PER^ and 
`r t.test_mouse.av.IFD_ABS_MUSC$p.value`  `r t.test_mouse.av.IFD_ABS_MOL$p.value` for IFD^raw^).

@@

#M2.raw
Q2.M2glm_IFD.raw <- glm(mean_IFD.2CO_ABS ~ strain, data= mouse.avs_4MM %>% filter(sex == "male"))
Sum.Q2.M2glm_IFD.raw <- summary(Q2.M2glm_IFD.raw)

@@

```{r Q2.IFD_rate.shortIFD, echo=FALSE, warning = FALSE, eval=FALSE}

@@
